---
title: "Assignment 01"
author: "Leo Camino | ltc31"
format: html
editor: visual
bibliography: references.bib
---

# First section: Demographic analysis of BioGrad Students - 2025

## Un-ordered list with at least two levels:

In the following list, I will present the names of students by country from the Cohort 25' that joined Duke for the [Biology Departmen at Duke](https://biology.duke.edu/):

### Biology Cohort 25' Students by country

-   United States of America 🇺🇸

    -   Charlotte Meigs

    -   Zöe Dietrich

    -   Brendan Lan

-   China 🇨🇳

    -   Yuzhe Yuan

    -   Alex Pei

    -   Chang Liu

-   India 🇮🇳

    -   Sharan Sekar

    -   Shreya Sarkar

-   Ecuador 🇪🇨

    -   Leo Camino

Summary table:

| Name            | Country                  | Cohort |
|-----------------|--------------------------|--------|
| Charlotte Meigs | United States of America | 25'    |
| Zöe Dietrich    | United States of America | 25'    |
| Brendan Lan     | United States of America | 25'    |
| Yuzhe Yuan      | China                    | 25'    |
| Alex Pei        | China                    | 25'    |
| Chang Liu       | China                    | 25'    |
| Sharan Sekar    | India                    | 25'    |
| Shreya Sarkar   | India                    | 25'    |
| Leo Camino      | Ecuador                  | 25'    |
| Yuzhe Yuan      | China                    | 25'    |

## Ordered list with at least two levels:

Here I provide a second list that presents the Cohort 25' students numbered by continent.

### Biology Cohort 25' Students by continent

1.  The Americas

    1.  Charlotte Meigs

    2.  Zöe Dietrich

    3.  Brendan Lan

    4.  Leo Camino

2.  Asia

    1.  Yuzhe Yuan

    2.  Alex Pei

    3.  Chang Liu

    4.  Sharan Sekar

    5.  Shreya Sarkar

# Second section: A random interesting text

Here I am writing this with the purpose of learning how to use *Italic,* **Bold**, and `verbatim` texts. Even though, I am proficient with the use of *Italic*, which involves only pressing ***"control"*** and ***"I"*** keys on Windows, I have no idea on how `monospaced texts work.` I have also used extensively **bold** words since I have worked a lot with text editors, such as Microsoft Word or Adobe InDesign. I guess `verbatim texts` are useful to explain something in text that involves sections of a code, for example:

*"The following code says: `sqrt(6) + 6!`."*

This could be specially useful for me, in the context of generating code for reviewers that are not proficient with structural coloration Data Analysis. I often have to use spectral analysis tools in R such as as Pavo2 [^1]. Being able to describe them on text with verbatim characters will give very important for me.

[^1]: Pavo2 Original paper:

    <https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.13174>

# Third section: Numerical Operations

## Section 3.2. Arithmetic Operations

```{r}

10 + 2 
```

**Addition:** I think this doesn't require a big explanation. One of the most basic arithmetic operations. A sum of two numbers adds both of them. Very romantic \<3.

```{r}

10 - 2
```

**Subtraction:** sometimes relationships don’t work out. Ten decided it’s better off without two, leaving it with 8.

```{r}

10 * 2
```

**Multiplication:** ten got ambitious, doubled its efforts, and became 20.

```{r}
10 / 2 
```

**Division:** ten decided to share equally with two. Both get 5. Sharing is caring!

```{r}
10 ^ 2
```

**Exponentiation:** ten flexed its power by raising itself to 2. That’s 100. Pure muscle

```{r}
10 ** 2
```

**Double asterisks** do the same as above. Ten still ends up at 100. Just another way to exponentiation. I am wondering if they made this for people that were unable to find the "\^" ASCII Code.

```{r}
10 %% 3 

```

**Modulo:** ten tried dividing by three, but three goes in only three times… leaving a lonely 1. That is the one we see here!

```{r}
10 %/% 3
```

**Integer division:** ten split itself into threes as many whole times as possible. It managed 3 times. The decimal wen somewhere, but it is not present in this code anymore hehe.

## Section 3.3. Complex Arithmetic Operations

```{r}
(10 + 2)/4-5  
```

The way this works is:

-   `a =``(10 + 2)`

-   `b = 4`

-   `c = 5`

We have the following structure: **`a/(b - c).`**

For this reason, 12/4 = 3. This means a/b = 3.

Now, to solve a/b - c we have to replace like = 3 - (5), which is equal to -2. Perfect sense.

```{r}
(10 + 2)/(4-5) 
```

In this case, we have a different format.

`a = (10 + 2)`

`b = (4 - 5)`

We are asked to solve a/b here. Solving all the parentheses first:

a = 12; b = -1.

12/(-1) = -12.

I think this makes perfect sense and shows why parentheses syntax must be correct in order to achieve the desired solution.

## Section 3.4. Other Numerical Operations

```{r}
sqrt(10)
```

`sqrt(10)` computes the square root of 10.

```{r}
10 ^ 0.5
```

`10 ^ 0.5` is another way of writing the square root of 10. It gives the same result as `sqrt(10)`.

```{r}
log(10)
```

`log(10)` returns the natural logarithm of 10 (base *e*).

```{r}
log10(10)
```

`log10(10)` returns the logarithm of 10 with base 10.

```{r}
exp(1)
```

`exp(1)` calculates *e* raised to the power of 1, which equals the constant *e* (\~2.718).

```{r}
abs(-10)
```

`abs(-10)` returns the absolute value of -10, which is 10.

```{r}
sin(3.14)
```

`sin(3.14)` calculates the sine of 3.14. Since 3.14 approximates π, the result is close to 0.

```{r}
cos(3.14)
```

`cos(3.14)` calculates the cosine of 3.14, which is close to -1 because 3.14 approximates π.

```{r}
pi
```

I think `pi` explains itself: pi = 3.1416...

```{r}
sin(pi)
```

`sin(pi)` returns a value very close to 0. I guess because of decimal computation it calculates a value close to zero, but not zero.

```{r}
cos(pi)
```

`cos(pi)` returns exactly -1, I guess somehow the decimals in this case didn't affect the final result.

## Section 3.5. Irrational numbers

```{r}
1/0

```

`1/0` evaluates to positive infinity (`Inf`) in R.\
This represents a number growing beyond any finite bound.

```{r}
-1/0

```

`-1/0` evaluates to negative infinity (`-Inf`).\
It indicates a value decreasing without bound.

```{r}
0/0
```

`0/0` is undefined mathematically. It returns `NaN` (Not a Number) to represent an indeterminate or invalid result.

## Section 3.6. Binary Relational Operators 

```{r}
10 > 2
```

10 is bigger than 2, therefore it is true.

```{r}
10 < 2
```

10 is not smaller than 2, therefore it is false.

```{r}
10 < (2 * 5)
```

10 is not smaller than ten (which is the result between the parentheses). It is False!

```{r}
10 <= (2 * 5)
```

Same as before, but considering the symbol is "smaller or equal", in this case it is correct, because 10 = 10.

```{r}
10 >= (3 * 3)
```

This is also correct, because 10 is bigger than 9.

```{r}
10 == (2 * 5)
```

The expression checks whether 10 is equal to 2 \* 5.\
Since 2 \* 5 evaluates to 10, the comparison is true.

```{r}
10 != (2 * 6)
```

The expression checks whether 10 is not equal to 2 \* 6.
Since 2 \* 6 equals 12, and 10 is not equal to 12, the result is true.

## Section 3.8. Surprising results

```{r}
sqrt(4)
```

The square root of 4 is 2. Not surprising if you are in 2025. Maybe very surprising around IV century.

```{r}
sqrt(10)
```

The square root of 10 is an irrational number, approximately 3.162278.

```{r}

sqrt(4)^2
```

Squaring the square root of 4 gives back 4.

```{r}
sqrt(10)^2
```

Same as the last result.

```{r}
sqrt(4)^2 == 4
```

This comparison returns true because squaring the square root of 4 gives exactly 4

```{r}
sqrt(10)^2 == 10
```

This is very confusing. It may return false because of small floating-point errors. Computers have a hard time representing irrational numbers.

```{r}
all.equal(sqrt(10)^2, 10) # look up help for all.equal
```

all.equal() checks if two numbers are nearly equal within a small tolerance. Here it returns TRUE, showing that sqrt(10)\^2 is effectively equal to 10 despite tiny rounding errors.
