---
title: "Problem Set 17"
author: "Leo Camino, ltc31"
date: today
date-format: iso
format: 
  html:
    embed-resources: true
editor: source
---

## **Problems**

1.  Show how to use `grep` to calculate the number of times the Ste12 DNA binding motif appears in the upstream region of the **first gene (YAR009C) in the table**. Account for variable positions in your count and make your regex as concise and readable as possible. Also make sure to deal with case!

    ```{r}
    #I found this section a little bit challenging already. 


    #awk -F'\t' 'BEGIN{IGNORECASE=1} $1=="YAR009C"{print toupper($3)}' STE12_promoter_table.tsv \ grep -oE '[TA][GA]AAAC[AGT]' wc -l

    #This is the line I used to account for variable positions. This returns "2"
    ```

2.  Write an Awk script, `uppercase_motif.awk`, that returns the 1st (systematic ID) and 3rd (nucleotide sequence) columns of an input file formatted like `STE12_promoter_table.csv` with the text in the third column converted to upper case.

    ```{r}

    #This is the following script I produced:

    # uppercase_motif.awk
    # Usage:
    #   awk -f uppercase_motif.awk -F'\t' STE12_promoter_table.tsv

    #BEGIN { FS = OFS = "\t" }

    #NR == 1 {
        # If there's a header, pass through a two-col header; else continue normally
    #    if ($1 ~ /systematic/i && $3 ~ /(sequence|upstream)/i) {
    #        print $1, $3
    #        next
    #    }
    #}
    #{
    #    print $1, toupper($3)
    #}

    ```

    I also run the following command:

    ```{r}
    #awk -f uppercase_motif.awk -F'\t' STE12_promoter_table.tsv > uppercase_output.tsv
    ```

    And finally I confirmed I had the new file `uppercase_output.tsv`

3.  Write an Awk script, `count_motif.awk`, that calculates the number of Ste12 DNA binding motifs found in the upstream region of each gene in `STE12_promoter_table.tsv`. Account for variable positions in your count; also make sure to deal with case!

    This is the script I produced for count_motif.awk.

    ```{r}

    # count_motif.awk
    # Usage:
    #   awk -f count_motif.awk -F'\t' STE12_promoter_table.tsv

    #BEGIN {
    #    FS = OFS = "\t"
    #    print "systematic_name", "motif_count"
    #    motif = "[TA][GA]AAAC[AGT]"
    #}
    #NR == 1 && $1 ~ /systematic/i && $3 ~ /(sequence|upstream)/i { next }

    #{
    #    seq = toupper($3)
    #    c = gsub(motif, "&", seq)   # gsub returns the number of matches
    #    print $1, c
    #}
    #``

    ```

    Your script should return a tab delimited table of the form:

    ```         
    systematic_name motif_count YAR009C          2 YBR067C          0 YCL027W          3 YDR365W-B        1 ...<output truncated>...
    ```

4.  Challenging: Write an Awk script, `motif_matches.awk`, that returns both the motif count and the found matches. Your output should return a tab-delimited table of the form:

    ```         
    systematic_name motif_count matches YAR009C 2   AGAAACT,aaaaacg, YBR067C 0    YCL027W 3   TGAAACA,TGAAACG,tgaaaca, YDR365W-B   1   tgaaacg,
    ```

    The Awk function `patsplit` (global substitution) can be used for splitting a string on a regex match. Read the [String Functions](https://www.gnu.org/software/gawk/manual/html_node/String-Functions.html) section of the [GNU Awk manual](https://www.gnu.org/software/gawk/manual/gawk.html) for more info. Hint: The `seps` array in the argument to `patsplit` is where you’ll find the specific matches found by your regex.

    I tried, but had several issues on this code:

```{r}

# motif_matches.awk

#BEGIN {
#    FS = OFS = "\t"
#    print "systematic_name", "motif_count", "matches"
#    motif = "[TA][GA]AAAC[AGT]"
#}
#NR == 1 && $1 ~ /systematic/i && $3 ~ /(sequence|upstream)/i { next }

#{
#    orig = $3
#    up   = toupper(orig)
#    matches = ""
#    count = 0
#    pos = 1

  
#    while (match(substr(up, pos), motif)) {
#        start = pos + RSTART - 1
#        mlen  = RLENGTH
#        m = substr(orig, start, mlen)
#        matches = matches m ","
#        count++
#        pos = start + 1   # allow overlaps (safe here)
#    }

#    print $1, count, matches
#}

```
